On the uniqueness and stability of dictionaries for sparse representation of noisy signals

### 2017-08-17 manuscripts/PNAS/review/point_by_point_response.txt

-----------------
Since the result is an extension of the noise-free result of (HS15) to a noisy case, the authors should provide better justification for the significance of contributions. The main result seems to be based on an approach similar to (HS15), and there is little to no discussion of the computability and tightness of the sufficient condition for stability in the main results. Thus, the submission does not seem to be greatly more valuable than (HS15) and also not conclusive regarding the central problems of sparse linear coding and dictionary learning. That being said, the paper (with the modifications below) would certainly merit publication in very solid journal like IEEE Transactions on Information Theory, where it would be a median paper, in my opinion.
-----------------

RESPONSE:

We agree with the reviewer that a straightforward extension of the approach and noise-free result of (HS15) to the noisy case — while still a significant advancement in applied mathematics with actual practical implications — would not necessarily merit publication in PNAS on its own. The reviewer, however, seems to have overlooked the fact that our main result (Thm. 1) generalizes the noise-free result of (HS15) far beyond a simple extension to the noisy case in ways which provide valuable insight into the dictionary learning model and that a significant deviation from the approach taken in (HS15) was necessary to prove these new facts. Moreover, our solution (Cor. 2) to the optimization formulation of the dictionary learning problem (Prob. 2) — that of most interest to optimization theorists and practitioners of their methods — is (to our knowledge) the first of its kind in both the noise-free and noisy domains. We discuss all of these points in more detail in our response to the reviewer’s later "detailed comments".

We have included in our updated manuscript a discussion of the computability of the sufficient conditions (they are, indeed, computable), as per the reviewer’s request. The main constants in our work -- C_1, C_2 -- are computable, elementarily, in terms of two positive quantities that are natural in our context: (a) the "Lower Bound" L_H(A) of a matrix A given a hypergraph H (the support sets of sparse x's generating the data — see *NOTE* below this paragraph), which is a measure of how distinguishable two x’s with supports in H are by their images under A, and (b) the value in Def. 4 of \xi(V_1,…,V_l) given subspaces V_i spanned by columns of A, which bounds the distance between intersections of such subspaces and each subspace individually. The first quantity (a) is, in some form or another, related to standard objects in the field and may be computed as the smallest singular value of a set of specific sub-matrices; in particular, for fixed sparsity constraint k on H, the lower bound L_H(A) is efficiently computable (polynomial time). Because deciding L_H(A) > 0 is already NP-hard (TP14), it seems likely that without fixing a sparsity level there is no efficient computation in general for (a). The calculation of (b) involves determining "canonical angles" between subspaces, which is usually reduced again to a numerical (efficient) singular value decomposition given the fixed subspaces.  However, upon close inspection, C_2 nonetheless requires an exponential number of instances of computing \xi. Thus, while our constants are effective (an algorithm exists to compute them), they are not necessarily polynomial-time computable.

### 2017-08-17 manuscripts/PNAS/chazthm_pnas.tex
chaz

Before moving on to our proofs, we discuss briefly the computability and tightness of our results. [* DO THIS *]
%\pagebreak

@@

\begin{definition}\label{SpecialSupportSet}\label{FriedrichsDefinition}
For subspaces $V_1, \ldots, V_\ell \subseteq \mathbb{R}^m$, set $r := 1$ when $\ell = 1$ and define for $\ell \geq 2$:
\begin{align*}
\xi(\{V_i\}_{i=1}^\ell) := 1 - \left(1 -  \max \prod_{i=1}^{\ell-1} \sin^2  \theta \left(V_i, \cap_{j>i} V_j \right)  \right)^{1/2},
\end{align*} 
%
where the maximum is taken over all orderings 
%\footnote{We modify the quantity in \cite{Deutsch12} in this way since the subspace ordering is irrelevant to our purpose.} 
of the $V_i$ and the angle $\theta \in (0,\frac{\pi}{2}]$ is defined implicitly as \cite[Def.~9.4]{Deutsch12}:
\begin{align*}
\cos{\theta(U,W)} := \max\left\{ |\langle \mathbf{u}, \mathbf{w} \rangle|: \substack{ \mathbf{u} \in U \cap (U \cap W)^\perp, \ \|\mathbf{u}\|_2 \leq 1 \\ \mathbf{w} \in W \cap (U \cap W)^\perp, \  \|\mathbf{w}\|_2 \leq 1 } \right\}.
\end{align*}
\end{definition}
Note that $\theta \in (0,\frac{\pi}{2}]$ implies $0 < r \leq 1$.\footnote{We acknowledge the counter-intuitive property that $\theta =  \pi/2$ when $U = V$ or $U \perp V$.}  
The constant $C_2$ in Lem.~\ref{MainLemma} can then be expressed as:  
\begin{align}\label{Cdef2}
C_2(\mathbf{A}, \mathcal{H}) := \frac{ 2^{|\mathcal{H}|} \max_{j \in [m]} \|\mathbf{A}_j\|_2}{ \min_{\mathcal{G} \subseteq \mathcal{H}} \xi( \{ \bm{\mathcal{A}}_S \}_{S \in \mathcal{G}}) },
\end{align}
%
which we remark is consistent with the assumption on $C_1$ in the proof of the case $k=1$ at the beginning of this section.

[* PUT COMPUTABILITY IN HERE *]

@@

% A significant implication is that the constant $C_1$ of Thm.~\ref{DeterministicUniquenessTheorem} can be computed efficiently if we restrict the class of possible supports for the $\mathbf{\bar x}_i$. 
In the last several decades, from JPEG's dominance in image compression to the discovery of intrinsic ``Gabor'' wavelets in natural images \cite{donoho2001can} to the detection of forgeries \cite{hughes2010, Olshausen10} to artificial ``cat detecting neurons'' in machine intelligence, sparse linear coding has become a standard tool in signal analysis, allowing for the unsupervised discovery of myriad insights into natural signals. 
%A compelling feature of this model is its simple instantiation of the principle of Occam's razor: a natural video, for instance, is modeled as a linear combination of a small number of spatiotemporal building blocks, each representing an archetypical feature latent in the data. 
In the field of theoretical neuroscience in particular, dictionary learning for sparse linear coding and related methods have recovered characteristic components of natural images \cite{Olshausen96, hurri1996image, bell1997independent, van1998independent} and sounds \cite{bellsejnowski1996, smithlewicki2006, Carlson12} that reproduce some response properties of certain cortical neurons. The motivation for our work here was to understand when such universal representations can be expected to emerge from the sparse linear coding model.

@@

One of the major innovations in our work is a theory of combinatorial designs for support sets (the ``singleton intersection property'') that provides a handle on the complexities of intersecting subspaces in dictionary learning problems. We incorporate this idea into a new fundamental lemma in combinatorial matrix theory (Lem.~\ref{MainLemma}) that will be useful in any noisy setting where uniqueness up to feature relabelling/scaling is desirable, and which draws upon the definition of a new matrix lower bound induced by a given hypergraph. Our novel combined combinatorial and analytical proof allows for the following new theoretical advances over previous works (e.g. \cite{Hillar15}) beyond an extension to the noisy case: 1) the number of dictionary elements can be unknown; 2) data require a polynomial, not exponential, number of distinct sparse supports and in certain settings a practical (polynomial) number of samples in total is sufficient; and 3) the spark condition is not necessary for recovery of the dictionary.

@@

In fact, uniqueness guarantees with minimal assumptions apply to all areas of data science and engineering that utilize learned sparse representations. For example, several groups have applied compressed sensing to signal processing tasks: MRI analysis \cite{lustig2008compressed}, image compression \cite{Duarte08}, and, more recently, the design of an ultrafast camera \cite{Gao14}. Given such effective uses of compressed sensing, it is only a matter of time before these systems incorporate dictionary learning to encode and process data (e.g., in a device that learns structure from motion \cite{kong2016prior}). In these cases, assurances such as those offered by our theorems certify that different devices (with different initialization, samples, etc.) will learn equivalent representations given enough data from statistically identical systems. To contrast with the current hot topic of ``Deep Learning'', there are few such uniqueness guarantees for these multi-layer feedforward models of data. Moreover, it is now well-known in the community that even minuscule amounts of (``adversarial'') signal noise can dramatically alter their output \cite{goodfellow2014explaining}. In particular, it appears that a main reason for the sustained interest in unsupervised (sparse) dictionary learning for data analysis has been the assumed well-posedness of the model, confirmation of which forms the core of our  findings.

>>>>

Before moving on to proofs, we discuss briefly the tightness of our results for quantities: (a) the required coding error $\varepsilon(\delta_1, \delta_2)$ sufficient to guarantee uniqueness in terms of desired dictionary and sparse representation errors $\delta_1, \delta_2$, and (b) the number of required samples, $N$.  With respect to (a), even with fixed dictionary, the recovery of sparse representations up to an error that is linear in the noise is classically optimal (up to constants).  %Thus, having both sufficient error in coding for uniqueness being linear in terms of both the desired dictionary and sparse vector recovery error is optimal.  
We regard (b) as a basic open problem: how many samples determine the sparse linear coding model?  If $k$ is held fixed or if the size of the support set of reconstructing codes is known to be polynomial in $\bar m$ and $k$, then a practical (polynomial) amount of data is enough to determine the model.\footnote{In the latter case, a reexamination of the pigeonholing argument in the proof of Thm.~\ref{DeterministicUniquenessTheorem} shows that only a polynomial number of samples distributed over a polynomial number of supports suffices.} Reasons to be skeptical that a polynomial amount suffice in general, however, can be found in \cite{tillmann2014computational, tillmann2015computational}.

@@

\begin{definition}\label{SpecialSupportSet}\label{FriedrichsDefinition}
For subspaces $V_1, \ldots, V_\ell \subseteq \mathbb{R}^m$, set $\xi := 1$ when $\ell = 1$ and define for $\ell \geq 2$:
\begin{align*}
\xi(\{V_i\}_{i=1}^\ell) := 1 - \left(1 -  \max \prod_{i=1}^{\ell-1} \sin^2  \theta \left(V_i, \cap_{j>i} V_j \right)  \right)^{1/2},
\end{align*} 
%
where the maximum is taken over all orderings 
%\footnote{We modify the quantity in \cite{Deutsch12} in this way since the subspace ordering is irrelevant to our purpose.} 
of the $V_i$ and the angle $\theta \in (0,\frac{\pi}{2}]$ is defined implicitly as \cite[Def.~9.4]{Deutsch12}:
\begin{align*}
\cos{\theta(U,W)} := \max\left\{ |\langle \mathbf{u}, \mathbf{w} \rangle|: \substack{ \mathbf{u} \in U \cap (U \cap W)^\perp, \ \|\mathbf{u}\|_2 \leq 1 \\ \mathbf{w} \in W \cap (U \cap W)^\perp, \  \|\mathbf{w}\|_2 \leq 1 } \right\}.
\end{align*}
\end{definition}
Note that $\theta \in (0,\frac{\pi}{2}]$ implies $0 < \xi \leq 1$.\footnote{We acknowledge the counter-intuitive property that $\theta =  \pi/2$ when $U = V$ or $U \perp V$.}  
The constant $C_2$ in Lem.~\ref{MainLemma} can then be expressed as:  
\begin{align}\label{Cdef2}
C_2(\mathbf{A}, \mathcal{H}) := \frac{ 2^{|\mathcal{H}|} \max_{j \in [m]} \|\mathbf{A}_j\|_2}{ \min_{\mathcal{G} \subseteq \mathcal{H}} \xi( \{ \bm{\mathcal{A}}_S \}_{S \in \mathcal{G}}) },
\end{align}
%
which we remark is consistent with the assumption on $C_1$ in the proof of the case $k=1$ at the beginning of this section.

Note that the explicit constants $C_1$ and $C_2$ are effectively computable. (The quantity $L_k$, may be computed as the smallest singular value of a certain matrix, while the quantity $\xi$ involves computing ``canonical angles'' between subspaces, which reduce again to an efficient singular value decomposition.) Because deciding $L_{k}(\mathbf{A}) > 0$ is already NP-hard \cite{tillmann2014computational}, it seems likely that without fixing a sparsity level there is no efficient computation $L_k$ in general; for fixed sparsity $k$, however, it is computable in polynomial time. Regardless, to calculate $C_2$ nonetheless requires an exponential number of queries to $\xi$. Thus, $C_1$ and $C_2$ are not necessarily polynomial-time computable.

@@

% A significant implication is that the constant $C_1$ of Thm.~\ref{DeterministicUniquenessTheorem} can be computed efficiently if we restrict the class of possible supports for the $\mathbf{\bar x}_i$. 
Over the last twenty years, from the discovery of intrinsic ``Gabor'' wavelets in natural images \cite{donoho2001can} to the detection of forgeries \cite{hughes2010, Olshausen10}, 
sparse linear coding has become a standard tool in signal analysis, allowing for the unsupervised discovery of myriad insights into natural signals in a variety of scientific fields. 
%A compelling feature of this model is its simple instantiation of the principle of Occam's razor: a natural video, for instance, is modeled as a linear combination of a small number of spatiotemporal building blocks, each representing an archetypical feature latent in the data. 
For theoretical neuroscience in particular, dictionary learning for sparse linear coding and related methods have recovered characteristic components of natural images \cite{Olshausen96, hurri1996image, bell1997independent, van1998independent} and sounds \cite{bellsejnowski1996, smithlewicki2006} that reproduce response properties of certain cortical neurons.  
In this work, we sought to understand when such universal representations be expected to emerge from the sparse linear coding model.
%to the recovery of ``mouse neuronal activity representing location on a track \cite{agarwal2014spatially}

@@

One of the major innovations in our work is a theory of combinatorial designs for support sets (the ``singleton intersection property'') key to the identification of the dictionary. We incorporate this idea into a new fundamental lemma in combinatorial matrix theory (Lem.~\ref{MainLemma}) that will be useful in any noisy setting where uniqueness up to feature relabelling/scaling is desirable, and which draws upon the definition of a new matrix lower bound induced by a given hypergraph. Our arguments allow for the following theoretical advances: the dictionary size can be unknown; data require a polynomial, not exponential, number of distinct sparse supports and in certain settings a polynomial number of samples in total is sufficient; and the spark condition is not necessary for dictionary recovery.

@@

In fact, uniqueness guarantees with minimal assumptions apply to all areas of data science and engineering that utilize learned sparse representations. For example, several groups have applied compressed sensing to signal processing tasks: MRI analysis \cite{lustig2008compressed}, image compression \cite{Duarte08}, and, more recently, the design of an ultrafast camera \cite{Gao14}. Given such effective uses of compressed sensing, it is only a matter of time before these systems incorporate dictionary learning to encode and process data (e.g., in a device that learns structure from motion \cite{kong2016prior}). In these cases, assurances such as those offered by our theorems certify that different devices (with different initialization, samples, etc.) will learn equivalent representations given enough data from statistically identical systems.

To contrast with the current hot topic of ``Deep Learning'', there are few such uniqueness guarantees for these multi-layer feedforward models of data. Moreover, there is evidence that even minuscule amounts of (``adversarial'') signal noise can dramatically alter their output \cite{goodfellow2014explaining}. In particular, a main reason for the sustained interest in dictionary learning for data analysis appears to be the assumed well-posedness of the model, confirmation of which forms the core of our  findings.
